---
title: "Common Integration Site (CIS) Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
params:
  is_dir: "05_insertions_repeatmask"
  connect_distance: 50000
  fdr_threshold: 0.05
---

<!--
Copyright 2025 Novartis Institutes for BioMedical Research Inc.

Licensed under the MIT License (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

https://www.mit.edu/~amini/LICENSE.md

Unless required by applicable law or agreed to in writing, software
distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and
limitations under the License.
-->

# Overview

This notebook identifies **Common Integration Sites (CIS)** — genomic regions
with statistically recurrent AAV integrations — using a position-based sliding
window approach with a data-driven minimum order threshold.

CIS are computed separately for each treatment group and ranked by statistical
significance using a Poisson test with per-chromosome normalization.

**Requirements:**

- R packages: `tidyverse`, `GenomicRanges`, `kableExtra`, `gt23`, `lisat`
- Input: `*_integration_sites.tsv` files from the TES-AAV pipeline

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, cache = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7
)

library(tidyverse)
library(GenomicRanges)
library(kableExtra)
library(gt23)
library(lisat)
```

# Data Loading & Preprocessing

```{r load-data}
# Discover integration site TSV files
is_files <- list.files(
  params$is_dir,
  pattern = "_integration_sites\\.tsv$",
  full.names = TRUE
)
stopifnot("No integration_sites.tsv files found" = length(is_files) > 0)

# Read and combine
is_data_all <- map_dfr(is_files, function(f) {
  df <- read_tsv(f, show_col_types = FALSE)
  df$sample <- sub("_integration_sites\\.tsv$", "", basename(f))
  df
})

cat(sprintf("Loaded %d integration sites across %d samples\n",
            nrow(is_data_all), n_distinct(is_data_all$sample)))
```

```{r filter-confidence}
# Exclude low-confidence sites (artifacts flagged by annotate_is_confidence.py)
is_data <- is_data_all %>%
  filter(confidence != "low" | is.na(confidence))

cat(sprintf("Retained %d / %d sites (excluded %d low-confidence artifacts)\n",
            nrow(is_data), nrow(is_data_all), nrow(is_data_all) - nrow(is_data)))
```

```{r sample-metadata}
# Parse sample metadata from file names
# Adapt this block to your sample naming convention.
# Expected pattern: {condition}_R1_A1_{donor}
is_data <- is_data %>%
  mutate(
    donor = str_extract(sample, "\\d{3}$"),
    condition = str_remove(sample, "_R1_A1_\\d{3}$"),
    treatment = case_when(
      str_detect(condition, "^unt")      ~ "AAV_only",
      str_detect(condition, "^ems2and3") ~ "EMS1and2_3weeks",
      str_detect(condition, "^ems2")     ~ "EMS_2weeks",
      str_detect(condition, "^ems3")     ~ "EMS_3weeks",
      TRUE ~ condition
    )
  )
```

# Gene Annotation (gt23)

Annotate integration sites with nearest gene, oncogene, and distance using
the `gt23` package.

```{r annotate-genes}
# Build GRanges per sample from cluster coordinates
is_for_annot <- is_data %>%
  mutate(
    first_chr = str_extract(cluster_chr, "^[^;]+"),
    first_chr = str_trim(first_chr),
    annot_strand = case_when(
      orientation == "fwd" ~ "+",
      orientation == "rev" ~ "-",
      TRUE ~ "*"
    )
  ) %>%
  filter(!is.na(first_chr), !is.na(cluster_start), !is.na(cluster_end))

gr_list <- split(is_for_annot, is_for_annot$sample) %>%
  lapply(function(df) {
    GRanges(
      seqnames = df$first_chr,
      ranges   = IRanges(start = df$cluster_start, end = df$cluster_end),
      strand   = df$annot_strand
    )
  })

# Annotate each sample
annotated_list <- list()
for (nm in names(gr_list)) {
  gr <- gr_list[[nm]]
  mcols(gr)$patient   <- "Sample1"
  mcols(gr)$refGenome <- "hg38"
  gr <- gt23::annotateIntSites(gr)
  annotated_list[[nm]] <- as.data.frame(gr)
}
annotated_data <- bind_rows(annotated_list, .id = "sample")

# Transfer annotations back
is_for_annot$nearestFeature     <- annotated_data$nearestFeature
is_for_annot$nearestFeatureDist <- annotated_data$nearestFeatureDist
is_for_annot$inFeature          <- annotated_data$inFeature
is_for_annot$nearestOncoFeature     <- annotated_data$nearestOncoFeature
is_for_annot$nearestOncoFeatureDist <- as.numeric(annotated_data$nearestOncoFeatureDist)

is_data <- is_data %>%
  left_join(
    is_for_annot %>% select(sample, site_id, nearestFeature, nearestFeatureDist,
                            inFeature, nearestOncoFeature, nearestOncoFeatureDist),
    by = c("sample", "site_id")
  )

cat(sprintf("Annotated %d / %d integration sites\n",
            sum(!is.na(is_data$nearestFeature)), nrow(is_data)))
```

# Common Integration Sites (CIS)

## Method

The minimum CIS order **K\*** is the smallest number of ISs in a window of size
`2 × connect_distance` such that fewer than 1 cluster of that size is expected
by chance anywhere in the genome (summing across all chromosomes, each with
its own size and IS count). This threshold is computed separately for each
treatment group.

A CIS region is called significant if it meets:

1. **Minimum order**: ≥K\* ISs in the region (data-driven, not arbitrary)
2. **Statistical significance**: FDR-adjusted p-value < `r params$fdr_threshold`
   (Poisson test with per-chromosome normalization)

Additional confidence flags (not used for filtering):

- **Multi-sample support**: ISs from ≥2 independent samples
- **Multi-donor support** (strongest evidence): ISs from both donors

```{r cis-prepare-data}
# Prepare lisat-format data
# Use upstream breakpoint coordinates; fall back to downstream if upstream is NA
IS_raw_lisat <- is_data %>%
  mutate(
    bp_chr = coalesce(chr_bp_upstream, chr_bp_downstream),
    bp_pos = coalesce(pos_bp_upstream, pos_bp_downstream)
  ) %>%
  filter(!is.na(bp_chr), !is.na(bp_pos)) %>%
  transmute(
    Sample = sample,
    SCount = count,
    Chr = sub("^chr", "", bp_chr),
    Locus = as.integer(bp_pos),
    nearest_gene_name = nearestFeature,
    treatment_group = treatment,
    donor = donor
  ) %>%
  filter(Chr %in% c(as.character(1:22), "X", "Y", "M")) %>%
  filter(!is.na(nearest_gene_name))

# Split into treatment groups
IS_aav_only <- IS_raw_lisat %>% filter(treatment_group == "AAV_only")
IS_ems      <- IS_raw_lisat %>% filter(treatment_group != "AAV_only")

cat(sprintf("AAV-only: %d ISs across %d samples\n",
            nrow(IS_aav_only), n_distinct(IS_aav_only$Sample)))
cat(sprintf("EMS-treated: %d ISs across %d samples\n",
            nrow(IS_ems), n_distinct(IS_ems$Sample)))
```

```{r cis-functions}
# Chromosome lengths (hg38)
chr_sizes_hg38 <- c(
  "1" = 248956422, "2" = 242193529, "3" = 198295559, "4" = 190214555,
  "5" = 181538259, "6" = 170805979, "7" = 159345973, "8" = 145138636,
  "9" = 138394717, "10" = 133797422, "11" = 135086622, "12" = 133275309,
  "13" = 114364328, "14" = 107043718, "15" = 101991189, "16" = 90338345,
  "17" = 83257441, "18" = 80373285, "19" = 58617616, "20" = 64444167,
  "21" = 46709983, "22" = 50818468, "X" = 156040895, "Y" = 57227415,
  "M" = 16569
)

# Compute data-driven minimum CIS order
#
# For each candidate order K (2, 3, ...), compute expected number of K-order
# clusters genome-wide under uniform random integration:
#   E[CIS_K] = sum_chr (L_i / W) * P(X >= K | Binom(n_i, W/L_i))
# K* = smallest K where E[CIS_K] < 1
compute_min_cis_order <- function(IS_df, connect_distance,
                                  chr_sizes = chr_sizes_hg38,
                                  significance = 1.0) {
  window_size <- 2 * connect_distance
  is_per_chr <- IS_df %>% count(Chr, name = "n_chr")

  for (K in 2:20) {
    expected_total <- 0
    for (i in seq_len(nrow(is_per_chr))) {
      chr <- is_per_chr$Chr[i]
      n_i <- is_per_chr$n_chr[i]
      L_i <- chr_sizes[chr]
      if (is.na(L_i) || n_i < K) next
      if (L_i <= window_size) next
      n_windows <- L_i / window_size
      p_k <- pbinom(K - 1, size = n_i, prob = window_size / L_i,
                     lower.tail = FALSE)
      expected_total <- expected_total + n_windows * p_k
    }
    if (expected_total < significance) {
      return(list(K_min = K, expected_clusters = expected_total))
    }
  }
  return(list(K_min = NA, expected_clusters = NA))
}

# Position-based sliding window CIS detection with per-chromosome Poisson test
compute_cis_sliding_window <- function(IS_df, connect_distance,
                                       chr_sizes = chr_sizes_hg38) {
  window_size <- 2 * connect_distance

  # Step 1: Data-driven minimum CIS order

  min_order <- compute_min_cis_order(IS_df, connect_distance, chr_sizes)
  K_min <- min_order$K_min
  cat(sprintf("Data-driven minimum CIS order: K* = %d\n", K_min))
  cat(sprintf("  (Expected K*-order clusters by chance: %.4f)\n",
              min_order$expected_clusters))

  # Report expected counts for orders below threshold
  is_per_chr_tmp <- IS_df %>% count(Chr, name = "n_chr")
  for (k in 2:max(2, K_min)) {
    expected_k <- 0
    for (i in seq_len(nrow(is_per_chr_tmp))) {
      chr <- is_per_chr_tmp$Chr[i]
      n_i <- is_per_chr_tmp$n_chr[i]
      L_i <- chr_sizes[chr]
      if (is.na(L_i) || n_i < k || L_i <= window_size) next
      n_windows <- L_i / window_size
      p_k <- pbinom(k - 1, size = n_i, prob = window_size / L_i,
                     lower.tail = FALSE)
      expected_k <- expected_k + n_windows * p_k
    }
    cat(sprintf("  Order %d: %.2f clusters expected by chance\n", k, expected_k))
  }

  # Step 2: Count ISs per chromosome
  is_per_chr <- IS_df %>% count(Chr, name = "n_chr")
  IS_df <- IS_df %>%
    left_join(is_per_chr, by = "Chr") %>%
    mutate(chr_length = chr_sizes[Chr]) %>%
    arrange(Chr, Locus)

  # Step 3: Count neighbors within ±connect_distance per IS
  cis_per_is <- IS_df %>%
    group_by(Chr) %>%
    mutate(
      n_neighbors = sapply(Locus, function(pos) {
        sum(abs(Locus - pos) <= connect_distance) - 1L
      }),
      lambda_chr = (n_chr - 1) * (window_size / chr_length)
    ) %>%
    ungroup()

  # Step 4: Keep only ISs meeting minimum order
  cis_hits <- cis_per_is %>% filter(n_neighbors >= K_min)

  if (nrow(cis_hits) == 0) {
    cat("No CIS regions meet the minimum order threshold.\n")
    return(tibble(
      cis_id = integer(), Chr = character(), start = numeric(), end = numeric(),
      n_is = integer(), n_unique_loci = integer(), n_samples = integer(),
      n_donors = integer(), span_bp = numeric(), genes = character(),
      chr_length = numeric(), n_is_on_chr = integer(),
      effective_window = numeric(), lambda_region = numeric(),
      p_value = numeric(), p_adj = numeric(), rank = integer(),
      K_min = integer()
    ))
  }

  # Step 5: Merge overlapping windows into CIS regions
  cis_gr <- GRanges(
    seqnames = paste0("chr", cis_hits$Chr),
    ranges = IRanges(
      start = pmax(1L, as.integer(cis_hits$Locus - connect_distance)),
      end   = as.integer(cis_hits$Locus + connect_distance)
    )
  )
  merged <- reduce(cis_gr)

  # Assign each IS to a merged CIS region
  all_gr <- GRanges(
    seqnames = paste0("chr", IS_df$Chr),
    ranges = IRanges(start = IS_df$Locus, end = IS_df$Locus)
  )
  olaps <- findOverlaps(all_gr, merged)
  IS_df$cis_id <- NA_integer_
  IS_df$cis_id[queryHits(olaps)] <- subjectHits(olaps)

  # Step 6: Summarize CIS regions with per-chromosome Poisson test
  cis_summary <- IS_df %>%
    filter(!is.na(cis_id)) %>%
    group_by(cis_id) %>%
    summarise(
      Chr = dplyr::first(Chr),
      start = min(Locus),
      end = max(Locus),
      n_is = n(),
      n_unique_loci = n_distinct(Locus),
      n_samples = n_distinct(Sample),
      n_donors = n_distinct(donor),
      span_bp = max(Locus) - min(Locus),
      genes = paste(sort(unique(nearest_gene_name)), collapse = "; "),
      samples_list = paste(sort(unique(Sample)), collapse = "; "),
      chr_length = dplyr::first(chr_length),
      n_is_on_chr = dplyr::first(n_chr),
      .groups = "drop"
    ) %>%
    filter(n_is >= K_min) %>%
    mutate(
      effective_window = pmin(span_bp + window_size, chr_length),
      lambda_region = (n_is_on_chr - 1) * (effective_window / chr_length),
      p_value = ppois(n_is - 1, lambda = lambda_region, lower.tail = FALSE),
      p_adj = p.adjust(p_value, method = "BH"),
      rank = rank(p_adj, ties.method = "min"),
      K_min = K_min
    ) %>%
    arrange(rank)

  return(cis_summary)
}
```

## lisat CIS Detection

The `lisat` package provides an independent CIS calling method for comparison.

### AAV-Only Group

```{r cis-lisat-aav, fig.width=10, fig.height=7}
IS_aav_cis_input <- IS_aav_only %>%
  select(Sample, SCount, Chr, Locus, nearest_gene_name)

CIS_top_aav <- tryCatch(
  CIS(IS_raw = IS_aav_cis_input, connect_distance = params$connect_distance),
  error = function(e) { message("lisat CIS (AAV-only): ", e$message); NULL }
)

if (!is.null(CIS_top_aav)) {
  CIS_overlap_aav <- CIS_overlap(CIS_data = CIS_top_aav, IS_raw = IS_aav_cis_input)
  CIS_overlap_aav
}
```

### EMS-Treated Group

```{r cis-lisat-ems, fig.width=10, fig.height=7}
IS_ems_cis_input <- IS_ems %>%
  select(Sample, SCount, Chr, Locus, nearest_gene_name)

CIS_top_ems <- tryCatch(
  CIS(IS_raw = IS_ems_cis_input, connect_distance = params$connect_distance),
  error = function(e) { message("lisat CIS (EMS): ", e$message); NULL }
)

if (!is.null(CIS_top_ems)) {
  CIS_overlap_ems <- CIS_overlap(CIS_data = CIS_top_ems, IS_raw = IS_ems_cis_input)
  CIS_overlap_ems
}
```

## Statistical CIS Ranking (Sliding Window + Poisson Test)

### AAV-Only Group

```{r cis-sig-aav}
cis_sig_aav <- compute_cis_sliding_window(IS_aav_only,
                                           connect_distance = params$connect_distance)

cis_sig_aav %>%
  head(30) %>%
  select(CIS = cis_id, Chr, Start = start, End = end,
         `ISs in region` = n_is, `Unique loci` = n_unique_loci,
         Samples = n_samples, Donors = n_donors,
         `Span (bp)` = span_bp, Genes = genes,
         `p-value` = p_value, `FDR` = p_adj, Rank = rank) %>%
  kbl(digits = 6,
      caption = paste0("AAV-only: CIS regions (min order K*=",
                       cis_sig_aav$K_min[1],
                       ", per-chr Poisson test, FDR-corrected)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")
```

### EMS-Treated Group

```{r cis-sig-ems}
cis_sig_ems <- compute_cis_sliding_window(IS_ems,
                                           connect_distance = params$connect_distance)

cis_sig_ems %>%
  head(30) %>%
  select(CIS = cis_id, Chr, Start = start, End = end,
         `ISs in region` = n_is, `Unique loci` = n_unique_loci,
         Samples = n_samples, Donors = n_donors,
         `Span (bp)` = span_bp, Genes = genes,
         `p-value` = p_value, `FDR` = p_adj, Rank = rank) %>%
  kbl(digits = 6,
      caption = paste0("EMS-treated: CIS regions (min order K*=",
                       cis_sig_ems$K_min[1],
                       ", per-chr Poisson test, FDR-corrected)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")
```

## CIS Comparison: AAV-only vs EMS-treated

```{r cis-comparison}
sig_threshold <- params$fdr_threshold

cis_aav_sig <- cis_sig_aav %>% filter(p_adj < sig_threshold) %>%
  mutate(group = "AAV_only")
cis_ems_sig <- cis_sig_ems %>% filter(p_adj < sig_threshold) %>%
  mutate(group = "EMS_treated")

cat(sprintf("Significant CIS regions (FDR < %.2f):\n", sig_threshold))
cat(sprintf("  AAV-only:    %d regions\n", nrow(cis_aav_sig)))
cat(sprintf("  EMS-treated: %d regions\n", nrow(cis_ems_sig)))

# Check for overlapping CIS regions between groups
if (nrow(cis_aav_sig) > 0 & nrow(cis_ems_sig) > 0) {
  gr_aav <- GRanges(seqnames = cis_aav_sig$Chr,
                    ranges = IRanges(start = cis_aav_sig$start, end = cis_aav_sig$end))
  gr_ems <- GRanges(seqnames = cis_ems_sig$Chr,
                    ranges = IRanges(start = cis_ems_sig$start, end = cis_ems_sig$end))
  shared_olaps <- findOverlaps(gr_aav, gr_ems, maxgap = params$connect_distance)
  cat(sprintf("  Shared (overlapping within %d bp): %d region pairs\n",
              params$connect_distance, length(shared_olaps)))
}

# Combined table
cis_all_sig <- bind_rows(cis_aav_sig, cis_ems_sig) %>%
  select(Group = group, Chr, Start = start, End = end,
         `ISs` = n_is, Samples = n_samples, Donors = n_donors,
         Genes = genes, FDR = p_adj) %>%
  arrange(FDR)

cis_all_sig %>%
  kbl(digits = 6,
      caption = sprintf("All significant CIS regions (FDR < %.2f)", sig_threshold)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")
```

## Significant CIS with Confidence Flags

```{r cis-valid-summary}
valid_cis_aav <- cis_sig_aav %>%
  filter(p_adj < sig_threshold) %>%
  mutate(group = "AAV_only",
         multi_sample = n_samples >= 2,
         multi_donor = n_donors >= 2)
valid_cis_ems <- cis_sig_ems %>%
  filter(p_adj < sig_threshold) %>%
  mutate(group = "EMS_treated",
         multi_sample = n_samples >= 2,
         multi_donor = n_donors >= 2)

valid_cis_all <- bind_rows(valid_cis_aav, valid_cis_ems)

if (nrow(valid_cis_all) > 0) {
  valid_cis_all %>%
    select(Chr, Start = start, End = end, Group = group,
           `ISs` = n_is, `Unique loci` = n_unique_loci,
           Samples = n_samples, Donors = n_donors,
           `Multi-sample` = multi_sample, `Multi-donor` = multi_donor,
           `Span (bp)` = span_bp, Genes = genes,
           FDR = p_adj, Rank = rank) %>%
    arrange(FDR) %>%
    kbl(digits = 6,
        caption = "Significant CIS regions with confidence flags") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
} else {
  cat("No CIS regions reach significance.\n")
}
```

## Integration Sites within Significant CIS Regions

Extract individual integration sites falling within significant CIS regions.

```{r cis-sites-extraction}
cis_ems_significant <- cis_sig_ems %>% filter(p_adj < sig_threshold)

# Prepare is_data with Chr/Locus columns for region queries
is_data_for_cis <- is_data %>%
  mutate(
    bp_chr_raw = coalesce(chr_bp_upstream, chr_bp_downstream),
    bp_pos_raw = coalesce(pos_bp_upstream, pos_bp_downstream),
    Chr = sub("^chr", "", bp_chr_raw),
    Locus = as.integer(bp_pos_raw)
  ) %>%
  filter(!is.na(Chr), !is.na(Locus),
         Chr %in% c(as.character(1:22), "X", "Y", "M"))

# Display ISs within each significant CIS region
if (nrow(cis_ems_significant) > 0) {
  for (i in seq_len(nrow(cis_ems_significant))) {
    cis_row <- cis_ems_significant[i, ]
    cis_start <- cis_row$start - params$connect_distance
    cis_end   <- cis_row$end + params$connect_distance

    is_in_cis <- is_data_for_cis %>%
      filter(Chr == cis_row$Chr,
             Locus >= cis_start,
             Locus <= cis_end)

    cat(sprintf("\n### CIS chr%s (%.3f–%.3f Mb): %d ISs from %d samples\n",
                cis_row$Chr, cis_row$start / 1e6, cis_row$end / 1e6,
                nrow(is_in_cis), n_distinct(is_in_cis$sample)))
    cat(sprintf("  AAV-only: %d | EMS-treated: %d\n",
                sum(is_in_cis$treatment == "AAV_only"),
                sum(is_in_cis$treatment != "AAV_only")))

    if (nrow(is_in_cis) > 0) {
      print(
        is_in_cis %>%
          arrange(Locus) %>%
          select(Sample = sample, Donor = donor, Chr, Position = Locus,
                 Reads = count, `Nearest Gene` = nearestFeature,
                 `Gene Dist` = nearestFeatureDist, `In Gene` = inFeature,
                 `Nearest Oncogene` = nearestOncoFeature,
                 `Oncogene Dist` = nearestOncoFeatureDist,
                 Orientation = orientation, Translocation = is_translocation,
                 Confidence = confidence, `BP Evidence` = bp_evidence) %>%
          kbl(caption = sprintf("ISs within CIS chr%s (%.3f–%.3f Mb ± %d kbp)",
                                cis_row$Chr, cis_row$start / 1e6,
                                cis_row$end / 1e6,
                                params$connect_distance / 1000)) %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                        font_size = 11) %>%
          scroll_box(height = "400px", width = "100%")
      )
    }
  }
}
```

# Session Info

```{r}
sessionInfo()
```
